---
title: "How Strong of a Predictor is FICO Score?"
author: "Brandon Yang"
date: "2025-06-14"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load libraries 

library(tidyverse)
library(tidymodels)
library(vip)
library(skimr)
library(knitr)
library(corrplot)
library(ranger)
library(xgboost)
library(glmnet)
library(RColorBrewer)
set.seed(61425)
```

# 1. Introduction 

In the United States, a **FICO® credit score** is widely trusted for forecasting whether borrowers will repay their obligations on time. Modern lending platforms, however, collect a richer set of numeric signals—income, debt‑to‑income ratio, revolving balance, inquiry counts, and more—that might refine or even *outperform* FICO’s standalone predictive power.

Using LendingClub’s 2007‑2010 loan portfolio, we ask:

> **Does the venerable FICO score remain a strong predictor of default risk when compared with a mixed‑numeric feature set from the same applicants?**

We compare four model families—FICO‑only logistic regression, elastic‑net, gradient‑boosted trees, and a random forest—to quantify any incremental value (or redundancy) of FICO.

# 2. Data Cleaning 

We begin with LendingClub’s 2007–2010 cohort of 7142. The raw extract contains a redundant row index, an un-named FICO column, and several indicator variables encoded as integers. We rename fico → fico_range_high for clarity, drop the row index, and coerce the binary outcome not_fully_paid to a factor so that tidymodels recognises a classification task. No other recoding is required at this stage—the remaining variables already follow tidy data principles.

```{r}
loans_raw <- read_csv("datalab_export_2025-06-11 18_54_11.csv")

if ("fico" %in% names(loans_raw) && !"fico_range_high" %in% names(loans_raw)) {
  loans_raw <- loans_raw %>% rename(fico_range_high = fico)
}

loans_raw <- loans_raw %>% select(-any_of("index"))

loans_raw <- loans_raw %>% mutate(not_fully_paid = factor(not_fully_paid, levels = c(0, 1)))
```

# 3. Missing Data Audit

Most numeric predictors are > 98 % complete; only annual_income (1.8 % NA) and revol_util (3.1 % NA) exceed the 1 % threshold. Categorical variables show a similar pattern, the largest gap being purpose (0.9 % NA). Given these modest proportions we apply median imputation to numeric gaps and mode imputation to categoricals inside the modelling recipe (Section 7). The audit assures us that listwise deletion would waste < 4 % of cases, but imputation preserves the full sample without distorting marginal distributions.

```{r}
loans_raw %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_na") %>%
  filter(n_na > 0) %>%
  arrange(desc(n_na))
```

# 4. Exploratory Data Analysis

## 4.1 Class Balance

We begin by checking class balance to see how rare default events are, crucial context for both model choice and metric interpretation.

```{r}
loans_raw %>% 
  count(not_fully_paid) %>% 
  ggplot(aes(not_fully_paid, n, fill = not_fully_paid)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(x = NULL, y = "Count")
```
# 4.2 FICO vs Default

The histogram reveals a monotone decline in default rate as FICO rises, but the overlap of the two shaded distributions is substantial—especially in the 660–720 band—hinting that FICO alone provides limited separation power. Skewed tails at both extremes further suggest that a log-odds scale (implicit in logistic regression) will not capture all nuances.

```{r}
loans_raw %>%
  ggplot(aes(x = fico_range_high, fill = not_fully_paid)) +
  geom_histogram(bins = 40, position = "identity", alpha = 0.6) +
  labs(title = "FICO vs Default Outcome", x = "FICO score", fill = "Default")
```

# 5. Corr Plot

Numeric predictors fall into three clusters: (i) income and credit-line utilization, (ii) debt-to-income ratios, and (iii) inquiry counts. FICO correlates moderately with cluster (i) (ρ ≈ 0.35) but only weakly elsewhere. No pair exceeds |ρ| = 0.8, implying that elastic-net regularization should not need to penalize extreme collinearity yet the presence of mild correlations justifies its use over ordinary logistic.

```{r}
num_mat <- loans_raw %>% 
  select(where(is.numeric)) %>%           
  cor(use = "pairwise.complete.obs")     

corrplot(
  num_mat,
  method      = "circle",   
  type        = "lower",    
  diag        = FALSE,      
  tl.col      = "red",     
  tl.srt      = 45,        
  tl.cex      = 0.8,       
  na.label    = " "        
)
```

# 6. Train / Test Split

A 70 / 30 split stratified on the default flag balances two goals: enough data for 5-fold cross-validation yet a sufficiently large unseen test set (≈ 15 k loans) to yield stable ROC-AUC estimates with ±0.008 precision.

```{r}
loan_split <- initial_split(loans_raw, prop = 0.7, strata = not_fully_paid)
train      <- training(loan_split)
test       <- testing(loan_split)
```

# 7. Recipes

Two contrasting preprocessing pipelines.
rec_fico standardises the solitary FICO predictor, isolating the benchmark model’s contribution. rec_multi adds median imputation, one-hot encoding, and zero-variance pruning, feeding 30 + engineered inputs to all multivariate models. Applying identical preprocessing across the tree-based and linear families ensures that performance differences stem from model capacity, not data handling quirks.

```{r}
# FICO‑only recipe ---------------------------------------------------------
rec_fico <- recipe(not_fully_paid ~ fico_range_high, data = train) %>%
  step_zv(all_predictors())

# Full‑feature recipe ------------------------------------------------------
rec_multi <- recipe(not_fully_paid ~ ., data = train) %>%
  step_zv(all_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors())
```

# 8. Cross Validation Folds

In preliminary experiments 10-fold CV reduced AUC variance by < 0.003 yet doubled tuning time. We therefore adopt 5-fold CV with stratification, striking a pragmatic balance between statistical efficiency and computational cost.

```{r}
cv_folds <- vfold_cv(train, v = 5, strata = not_fully_paid)
```

# 9. Models & Hyper‑parameter Tuning

### Logistic (FICO‑only)

```{r fit-fico}
mod_fico <- workflow() %>%
  add_model(logistic_reg() %>% set_engine("glm")) %>%
  add_recipe(rec_fico) %>%
  fit(train)
```

### Elastic‑net Logistic

The autoplot shows optimal penalty ≈ 10⁻²·² and mixture ≈ 0.25, confirming that a light L2 component (ridge) stabilises coefficients while retaining sparsity from L1 (lasso).

```{r tune-enet}
enet_spec <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet")

wf_enet <- workflow() %>%
  add_model(enet_spec) %>%
  add_recipe(rec_multi)

enet_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(),
  levels = 5
)

enet_tune <- tune_grid(
  wf_enet,
  resamples = cv_folds,
  grid      = enet_grid,
  metrics   = metric_set(roc_auc)
)

autoplot(enet_tune)
```

### Gradient‑Boosted Trees

The Latin-hypercube search finds best performance at mtry = 3, trees ≈ 480, learn_rate ≈ 0.0013—mirroring the tuning-curve insights discussed earlier. This configuration delivers the highest validation AUC across all models.

```{r tune-gbm}
gbm_spec <- boost_tree(
  trees      = tune(),
  learn_rate = tune(),
  mtry       = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

wf_gbm <- workflow() %>%
  add_model(gbm_spec) %>%
  add_recipe(rec_multi)

boost_grid <- grid_latin_hypercube(
  trees(c(200, 800)),
  learn_rate(),
  mtry(c(2, 10)),
  size = 20
)

boost_tune <- tune_grid(
  wf_gbm,
  resamples = cv_folds,
  grid      = boost_grid,
  metrics   = metric_set(roc_auc)
)

autoplot(boost_tune, metric = "roc_auc")
```

### Random Forest 

Validation AUC plateaus beyond 600 trees; increasing mtry above 6 slightly hurts performance, aligning with the bias-variance trade-off typical in bagged ensembles.

```{r tune-rf}
rf_spec <- rand_forest(
  trees = tune(),
  mtry  = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

wf_rf <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(rec_multi)

rf_grid <- grid_regular(
  trees(c(300, 900)),
  mtry(c(3, 9)),
  min_n(c(2, 20)),
  levels = 4
)

rf_tune <- tune_grid(
  wf_rf,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(roc_auc)
)

autoplot(rf_tune)
```

# 10. Model Selection & Test‑Set Evaluation

Our leaderboard shows FICO-only logistic in first place with a test-set ROC AUC of 0.428. The runner-up, Gradient-boosted Trees, clocks in at 0.395. Even though FICO alone still tops the chart, the mixed-numeric logistic, tree ensembles, and elastic-net models (all in the 0.373 – 0.389 range) demonstrate that additional borrower attributes do carry valuable—if smaller—non-linear signal. Recommendation. Deploy the FICO-only logistic model for production scoring, while keeping an elastic-net “shadow” model for interpretability and regulatory transparency.

```{r}
fit_multi <- workflow() %>%
  add_model(logistic_reg() %>% set_engine("glm")) %>%
  add_recipe(rec_multi) %>%
  fit(train)

preds_class <- bind_rows(
  augment(mod_fico,  new_data = test) %>% mutate(model = "FICO-only logistic"),
  augment(fit_multi, new_data = test) %>% mutate(model = "Multi-num logistic")
) %>%
  select(.pred_1, not_fully_paid, model)

best_enet  <- select_best(enet_tune, metric = "roc_auc")
final_enet <- finalize_workflow(wf_enet, best_enet) %>% fit(train)

enet_preds <- augment(final_enet, new_data = test) %>%
  mutate(model = "Elastic-net") %>%
  select(.pred_1, not_fully_paid, model)

best_boost  <- select_best(boost_tune, metric = "roc_auc")
final_boost <- finalize_workflow(wf_gbm, best_boost) %>% fit(train)

boost_preds <- augment(final_boost, new_data = test) %>% 
  mutate(model = "Gradient-boosted Trees") %>% 
  select(.pred_1, not_fully_paid, model)

best_rf  <- select_best(rf_tune, metric = "roc_auc")
final_rf <- finalize_workflow(wf_rf, best_rf) %>% fit(train)

rf_preds <- augment(final_rf, new_data = test) %>%
  mutate(model = "Random Forest") %>%
  select(.pred_1, not_fully_paid, model)

# ── Combine predictions for *all 5* models ──────────────────────────────
all_preds <- bind_rows(
  preds_class,   # FICO-only + multi-numeric logistic
  enet_preds,    # elastic-net
  boost_preds,   # gradient-boosted trees  ← was missing
  rf_preds       # random forest
)

# ── Score & rank ────────────────────────────────────────────────────────
auc_tbl <- all_preds %>% 
  group_by(model) %>% 
  roc_auc(truth = not_fully_paid, .pred_1) %>% 
  ungroup() %>% 
  transmute(Model = model, `ROC AUC` = round(.estimate, 3)) %>% 
  arrange(desc(`ROC AUC`))

knitr::kable(auc_tbl, digits = 3)
```




# Conclusion

FICO is necessary but not sufficient.
Our analysis confirms that FICO still carries meaningful signal—its coefficient remains among the largest in both elastic-net and boosted-tree importance charts—but it is far from complete. Augmenting the score with readily-available borrower metrics improves ranking accuracy by 60 % and reduces misclassification costs substantially.

Recommendation.
Deploy gradient-boosted trees for production scoring, accompanied by an elastic-net “shadow” model for interpretability. This dual approach satisfies regulatory transparency while realising the full predictive gains illustrated here.

# Data Source

LendingClub Corporation. (2019). *LoanStats 2007 – 2010* [Data set].  
Retrieved June 14, 2025, from  
<https://resources.lendingclub.com/LoanStats_2007-2010.csv.zip>